Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-27 paper/zhang-cook-laa-langrene-menendez.Rmd
Minor fixes in manuscript

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics 
    url: Melbourne, Australia
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    url: Vienna, Austria
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    url: Melbourne, Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  patricia.menendez@monash.edu 
abstract: A guided tour helps to visualise high-dimensional data by showing low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, might have multiple local maxima as well as an optimal maximum, and are constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package \pkg{ferrn}.
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{mathtools}
  \usepackage[ruled, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

The remainder of the paper is organised as follows. 
The next section provides an overview of optimisation methods, specifically random search and line search methods.  A review of the projection pursuit guided tour, an overview of the optimisation problem and outlines of three existing algorithms follows.
The third section presents the new visual diagnostics, including the design of a data structure to capture information during the optimisation, from which several diagnostic plots are created.
An illustration of how the diagnostic plots can be used to examine the performance of different optimisers and guide improvements to the existing algorithms applications is shown using simulated data.
Finally, an explanation of the implementation in the R package, \CRANpkg{ferrn} [@ferrn], is provided.

@@

Gradient-based methods are commonly used to optimise an objective function, with the most notable one being the gradient ascent (descent) method. Although these methods are popular, they rely on the availability of the objective function derivatives. As will be shown in the next section, the independent variables in our optimisation problem are the entries of a projection matrix and the computational time required to perform differentiation on a matrix could impede the rendering of tour animation. In addition, some objective functions rely on the empirical distribution of the data, which makes it in general not possible to get the gradient. Hence, gradient-based methods are not the focus of this paper and consideration will be given to derivative-free methods.

@@

\noindent where $f$ and $\mathbf{X}$ are defined as in the previous section, $\mathcal{A}$ is the set of all $p$-dimensional projection bases, $I_d$ is the $d$-dimensional identity matrix, and the constraint ensures the projection bases, $\mathbf{A}$, to be orthonormal. It is worth noticing the following: 1) The optimisation is constrained and the orthonormality constraint imposes a geometrical structure on the bases space: it forms a Stiefel manifold. 2) There may be index functions for which the objective function might not be differentiable. 3) While finding the global optimum is the goal of the optimisation problem, interesting projections may also appear in the local optimum. 4) The optimisation should be fast to compute since the tour animation is viewed by the users during the optimisation.

@@

Three optimisers have been implemented in the \pkg{tourr} [@tourr] package: creeping random search (CRS), simulated annealing (SA), and pseudo-derivative (PD). Creeping random search (CRS) is a random search optimiser that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is generated randomly. $\mathbf{A}_{l}$ is then orthonormalised to fulfil the basis constraint. If $\mathbf{A}_{l}$ has an index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the optimiser outputs $\mathbf{A}_{l}$ for guided tour to construct an interpolation path. The neighbourhood parameter $\alpha$ is adjusted by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$ before the next iteration starts. The optimiser terminates when the maximum number of iteration $l_{\max}$ is reached before a better basis can be found. The algorithm of CRS is summarised in Algorithm \ref{random-search} in the appendix. @posse1995projection has proposed a slightly different cooling scheme by introducing a halving parameter $c$. In his proposal, the $\alpha$ is only adjusted if the last iteration takes more than $c$ times to find a better basis.

Simulated annealing (SA) uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current one. Given an initial value of $T_0$, the "temperature" at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SA gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even if the candidate basis does not have a higher index value and hence enables the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} in the appendix highlights how SA differs from CRS in the inner loop.

Pseudo-derivative (PD) search uses a different strategy than CRS and SA. Rather than randomly sample the basis space, PD first computes a search direction by evaluating bases close to the current basis. A step size is then chosen along the corresponding geodesic by another optimisation over an 90 degree angle from $-\pi/4$ to $\pi/4$. The resulting candidate basis $\mathbf{A}_{**}$ is returned for the current iteration if it has a higher index value than the current one. Algorithm \ref{search-geodesic} in the appendix summarises the inner loop of the PD.

>>>>

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics 
    url: Melbourne, Australia
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    url: Vienna, Austria
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    url: Melbourne, Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    url: Melbourne, Australia
    email:  patricia.menendez@monash.edu 
abstract: A guided tour helps to visualise high-dimensional data by showing low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, might have multiple local maxima as well as an optimal maximum, and are constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package \pkg{ferrn}.
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{mathtools}
  \usepackage[ruled, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

The remainder of the paper is organised as follows. 
The next section provides an overview of optimisation methods, specifically random search and line search methods.  A review of the projection pursuit guided tour, an overview of the optimisation problem and outlines of three existing algorithms follows.
The third section presents the new visual diagnostics, including the design of a data structure to capture information during the optimisation process, from which several diagnostic plots are created.
An illustration of how the diagnostic plots can be used to examine the performance of different optimisers and guide improvements to existing algorithms is shown using simulated data.
Finally, an explanation of the implementation in the R package, \CRANpkg{ferrn} [@ferrn], is provided.

@@

Gradient-based methods are commonly used to optimise an objective function, with the most notable one being the gradient ascent (descent) method. Although these methods are popular, they rely on the availability of the objective function derivatives. As will be shown in the next section, the independent variables in our optimisation problem are the entries of a projection matrix, and the computational time required to perform differentiation on a matrix could impede the rendering of tour animation. In addition, some objective functions rely on the empirical distribution of the data, which makes it in general not possible to get the gradient. Hence, gradient-based methods are not the focus of this paper and consideration will be given to derivative-free methods.

@@

\noindent where $f$ and $\mathbf{X}$ are defined as in the previous section, $\mathcal{A}$ is the set of all $p$-dimensional projection bases, $I_d$ is the $d$-dimensional identity matrix, and the constraint ensures the projection bases, $\mathbf{A}$, to be orthonormal. It is worth noticing the following: 1) The optimisation is constrained and the orthonormality constraint imposes a geometrical structure on the bases space: it forms a Stiefel manifold. 2) There may be index functions for which the objective function might not be differentiable. 3) While finding the global optimum is the goal of the optimisation problem, interesting projections may also appear in the local optimum. 4) The optimisation should be fast to compute since the tour animation is viewed by the users during the optimisation process.

@@

Three optimisers have been implemented in the \pkg{tourr} [@tourr] package: creeping random search (CRS), simulated annealing (SA), and pseudo-derivative (PD). Creeping random search (CRS) is a random search optimiser that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha \in [0,1]$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is generated randomly. $\mathbf{A}_{l}$ is then orthonormalised to fulfil the basis constraint. If $\mathbf{A}_{l}$ has an index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the optimiser outputs $\mathbf{A}_{l}$ for guided tour to construct an interpolation path. The neighbourhood parameter $\alpha$ is adjusted by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$ before the next iteration starts. The optimiser terminates when the maximum number of iteration $l_{\max}$ is reached before a better basis can be found. The algorithm of CRS is summarised in Algorithm \ref{random-search} in the appendix. @posse1995projection has proposed a slightly different cooling scheme by introducing a halving parameter $c$. In his proposal $\alpha$ is only adjusted if the last iteration takes more than $c$ times to find a better basis.

Simulated annealing (SA) uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current one. Given an initial value of $T_0 \in \mathbb{R^{+}}$, the "temperature" at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SA gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)} \in \mathbb{R}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even if the candidate basis does not have a higher index value and hence enables the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} in the appendix highlights how SA differs from CRS in the inner loop.

Pseudo-derivative (PD) search uses a different strategy than CRS and SA. Rather than randomly sample the basis space, PD first computes a search direction by evaluating bases close to the current basis. A step size is then chosen along the corresponding geodesic by another optimisation over a 90 degree angle from $-\pi/4$ to $\pi/4$. The resulting candidate basis $\mathbf{A}_{**}$ is returned for the current iteration if it has a higher index value than the current one. Algorithm \ref{search-geodesic} in the appendix summarises the inner loop of the PD.

### 2021-03-27 paper/zhang-cook-laa-langrene-menendez.tex
Minor fixes in manuscript

\abstract{%
A guided tour helps to visualise high-dimensional data by showing
low-dimensional projections along a projection pursuit optimisation
path. Projection pursuit is a generalisation of principal component
analysis, in the sense that different indexes are used to define the
interestingness of the projected data. While much work has been done in
developing new indexes in the literature, less has been done on
understanding the optimisation. Index functions can be noisy, might have
multiple local maxima as well as an optimal maximum, and are constrained
to generate orthonormal projection frames, which complicates the
optimization. In addition, projection pursuit is primarily used for
exploratory data analysis, and finding the local maxima is also useful.
The guided tour is especially useful for exploration, because it
conducts geodesic interpolation connecting steps in the optimisation and
shows how the projected data changes as a maxima is approached. This
work provides new visual diagnostics for examining a choice of
optimisation procedure, based on the provision of a new data object
which collects information throughout the optimisation. It has helped to
diagnose and fix several problems with projection pursuit guided tour.
This work might be useful more broadly for diagnosing optimisers, and
comparing their performance. The diagnostics are implemented in the R
package \pkg{ferrn}.
}

@@

The remainder of the paper is organised as follows. The next section
provides an overview of optimisation methods, specifically random search
and line search methods. A review of the projection pursuit guided tour,
an overview of the optimisation problem and outlines of three existing
algorithms follows. The third section presents the new visual
diagnostics, including the design of a data structure to capture
information during the optimisation, from which several diagnostic plots
are created. An illustration of how the diagnostic plots can be used to
examine the performance of different optimisers and guide improvements
to the existing algorithms applications is shown using simulated data.
Finally, an explanation of the implementation in the R package,
\CRANpkg{ferrn} \citep{ferrn}, is provided.

@@

{\centering \includegraphics[width=0.7\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/tour-path}

@@

\noindent where \(f\) and \(\mathbf{X}\) are defined as in the previous
section, \(\mathcal{A}\) is the set of all \(p\)-dimensional projection
bases, \(I_d\) is the \(d\)-dimensional identity matrix, and the
constraint ensures the projection bases, \(\mathbf{A}\), to be
orthonormal. It is worth noticing the following: 1) The optimisation is
constrained and the orthonormality constraint imposes a geometrical
structure on the bases space: it forms a Stiefel manifold. 2) There may
be index functions for which the objective function might not be
differentiable. 3) While finding the global optimum is the goal of the
optimisation problem, interesting projections may also appear in the
local optimum. 4) The optimisation should be fast to compute since the
tour animation is viewed by the users during the optimisation.

@@

Three optimisers have been implemented in the \pkg{tourr} \citep{tourr}
package: creeping random search (CRS), simulated annealing (SA), and
pseudo-derivative (PD). Creeping random search (CRS) is a random search
optimiser that samples a candidate basis \(\mathbf{A}_{l}\) in the
neighbourhood of the current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is generated randomly. \(\mathbf{A}_{l}\)
is then orthonormalised to fulfil the basis constraint. If
\(\mathbf{A}_{l}\) has an index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the optimiser outputs \(\mathbf{A}_{l}\)
for guided tour to construct an interpolation path. The neighbourhood
parameter \(\alpha\) is adjusted by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next iteration
starts. The optimiser terminates when the maximum number of iteration
\(l_{\max}\) is reached before a better basis can be found. The
algorithm of CRS is summarised in Algorithm \ref{random-search} in the
appendix. \citet{posse1995projection} has proposed a slightly different
cooling scheme by introducing a halving parameter \(c\). In his
proposal, the \(\alpha\) is only adjusted if the last iteration takes
more than \(c\) times to find a better basis.

Simulated annealing (SA) uses the same sampling process as CRS but
allows a probabilistic acceptance of a basis with lower index value than
the current one. Given an initial value of \(T_0\), the ``temperature''
at iteration \(l\) is defined as \(T(l) = \frac{T_0}{\log(l + 1)}\).
When a candidate basis fails to have an index value larger than the
current basis, SA gives it a second chance to be accepted with
probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the optimiser to make a move and explore the basis
space even if the candidate basis does not have a higher index value and
hence enables the optimiser to jump out of a local optimum. The
algorithm \ref{simulated-annealing} in the appendix highlights how SA
differs from CRS in the inner loop.

Pseudo-derivative (PD) search uses a different strategy than CRS and SA.
Rather than randomly sample the basis space, PD first computes a search
direction by evaluating bases close to the current basis. A step size is
then chosen along the corresponding geodesic by another optimisation
over an 90 degree angle from \(-\pi/4\) to \(\pi/4\). The resulting
candidate basis \(\mathbf{A}_{**}\) is returned for the current
iteration if it has a higher index value than the current one. Algorithm
\ref{search-geodesic} in the appendix summarises the inner loop of the
PD.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-search-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-interp-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-pca-animated-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-tour-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-torus-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/polish-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/kol-result-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/flip-sign-1}

>>>>

\abstract{%
A guided tour helps to visualise high-dimensional data by showing
low-dimensional projections along a projection pursuit optimisation
path. Projection pursuit is a generalisation of principal component
analysis, in the sense that different indexes are used to define the
interestingness of the projected data. While much work has been done in
developing new indexes in the literature, less has been done on
understanding the optimisation. Index functions can be noisy, might have
multiple local maxima as well as an optimal maximum, and are constrained
to generate orthonormal projection frames, which complicates the
optimization. In addition, projection pursuit is primarily used for
exploratory data analysis. The guided tour is especially useful for
exploration, because it conducts geodesic interpolation connecting steps
in the optimisation and shows how the projected data changes as a maxima
is approached. This work provides new visual diagnostics for examining a
choice of optimisation procedure, based on the provision of a new data
object which collects information throughout the optimisation. It has
helped to diagnose and fix several problems with projection pursuit
guided tour. This work might be useful more broadly for diagnosing
optimisers, and comparing their performance. The diagnostics are
implemented in the R package \pkg{ferrn}.
}

@@

The remainder of the paper is organised as follows. The next section
provides an overview of optimisation methods, specifically random search
and line search methods. A review of the projection pursuit guided tour,
an overview of the optimisation problem and outlines of three existing
algorithms follows. The third section presents the new visual
diagnostics, including the design of a data structure to capture
information during the optimisation process, from which several
diagnostic plots are created. An illustration of how the diagnostic
plots can be used to examine the performance of different optimisers and
guide improvements to existing algorithms is shown using simulated data.
Finally, an explanation of the implementation in the R package,
\CRANpkg{ferrn} \citep{ferrn}, is provided.

@@

{\centering \includegraphics[width=0.7\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/tour-path}

@@

\noindent where \(f\) and \(\mathbf{X}\) are defined as in the previous
section, \(\mathcal{A}\) is the set of all \(p\)-dimensional projection
bases, \(I_d\) is the \(d\)-dimensional identity matrix, and the
constraint ensures the projection bases, \(\mathbf{A}\), to be
orthonormal. It is worth noticing the following: 1) The optimisation is
constrained and the orthonormality constraint imposes a geometrical
structure on the bases space: it forms a Stiefel manifold. 2) There may
be index functions for which the objective function might not be
differentiable. 3) While finding the global optimum is the goal of the
optimisation problem, interesting projections may also appear in the
local optimum. 4) The optimisation should be fast to compute since the
tour animation is viewed by the users during the optimisation process.

@@

Three optimisers have been implemented in the \pkg{tourr} \citep{tourr}
package: creeping random search (CRS), simulated annealing (SA), and
pseudo-derivative (PD). Creeping random search (CRS) is a random search
optimiser that samples a candidate basis \(\mathbf{A}_{l}\) in the
neighbourhood of the current basis \(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha \in [0,1]\) controls the radius of the sampling
neighbourhood and \(\mathbf{A}_{\text{rand}}\) is generated randomly.
\(\mathbf{A}_{l}\) is then orthonormalised to fulfil the basis
constraint. If \(\mathbf{A}_{l}\) has an index value higher than the
current basis \(\mathbf{A}_{\text{cur}}\), the optimiser outputs
\(\mathbf{A}_{l}\) for guided tour to construct an interpolation path.
The neighbourhood parameter \(\alpha\) is adjusted by a cooling
parameter: \(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next
iteration starts. The optimiser terminates when the maximum number of
iteration \(l_{\max}\) is reached before a better basis can be found.
The algorithm of CRS is summarised in Algorithm \ref{random-search} in
the appendix. \citet{posse1995projection} has proposed a slightly
different cooling scheme by introducing a halving parameter \(c\). In
his proposal \(\alpha\) is only adjusted if the last iteration takes
more than \(c\) times to find a better basis.

Simulated annealing (SA) uses the same sampling process as CRS but
allows a probabilistic acceptance of a basis with lower index value than
the current one. Given an initial value of \(T_0 \in \mathbb{R^{+}}\),
the ``temperature'' at iteration \(l\) is defined as
\(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have
an index value larger than the current basis, SA gives it a second
chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)} \in \mathbb{R}\) denotes the index value of a given
basis. This implementation allows the optimiser to make a move and
explore the basis space even if the candidate basis does not have a
higher index value and hence enables the optimiser to jump out of a
local optimum. The algorithm \ref{simulated-annealing} in the appendix
highlights how SA differs from CRS in the inner loop.

Pseudo-derivative (PD) search uses a different strategy than CRS and SA.
Rather than randomly sample the basis space, PD first computes a search
direction by evaluating bases close to the current basis. A step size is
then chosen along the corresponding geodesic by another optimisation
over a 90 degree angle from \(-\pi/4\) to \(\pi/4\). The resulting
candidate basis \(\mathbf{A}_{**}\) is returned for the current
iteration if it has a higher index value than the current one. Algorithm
\ref{search-geodesic} in the appendix summarises the inner loop of the
PD.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-search-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-interp-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-pca}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-pca-animated-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-tour-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/toy-torus-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/interruption-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/polish-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/noisy-better-geo-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/kol-result-1}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/pmen0008/Desktop/paper-tour-vis/figs/flip-sign-1}
